How do changes in ex vary between countries?

### 2022-05-05 draft_working_paper.Rmd
consider another summary graph

```{r}
compare_best_breakmodel_with_null %>% 
  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
  unnest(all_taus) %>% 
  ungroup() %>% left_join(
    segmented_breakpoint_models %>% 
      select(code, x, sex, bp_ftest_pval = ftest_pval, bp_breakpoint = breakpoint, bp_breakpoint_se = breakpoint_se)
  ) %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  mutate(
    ftest_cat = case_when(
      ftest_pval < 0.01 ~ "<0.01",
      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    ftest_cat = factor(ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  mutate(
    bp_ftest_cat = case_when(
      bp_ftest_pval < 0.01 ~ "<0.01",
      between(bp_ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(bp_ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    bp_ftest_cat = factor(bp_ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  ggplot(aes(tau, aic)) +
  geom_line() + 
  geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
  facet_wrap(~population, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "<0.01" = "darkred", 
      "0.01-0.05"  = "red", 
      "0.05-0.10" = "orange", 
      "Not Stat. Sig." = "lightblue"
      )
  ) + 
  geom_vline(
    aes(
      xintercept = bp_breakpoint,
      colour = bp_ftest_cat  
    )
  ) + 
  geom_rect(
    aes(
      xmin = bp_breakpoint - 1.96 * bp_breakpoint_se, 
      xmax = bp_breakpoint + 1.96 * bp_breakpoint_se
    ),
      ymin = -Inf, ymax = Inf,
    alpha = 0.01
  ) + 
  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
  labs(
    title = "Penalised fit (AIC) against proposed changepoint year, various population",
    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
    colour = "Statistical\nsignificance", 
    x = "Year proposed",
    y = "AIC (lower is better)",
    caption = "Vertical lines and bands show estimates and 95% CIs from segmented package"
  )
```

@@

>>>>

```{r}
compare_best_breakmodel_with_null %>% 
  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
  unnest(all_taus) %>% 
  ungroup() %>% left_join(
    segmented_breakpoint_models %>% 
      select(code, x, sex, bp_ftest_pval = ftest_pval, bp_breakpoint = breakpoint, bp_breakpoint_se = breakpoint_se)
  ) %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  mutate(
    ftest_cat = case_when(
      ftest_pval < 0.01 ~ "<0.01",
      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    ftest_cat = factor(ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  mutate(
    bp_ftest_cat = case_when(
      bp_ftest_pval < 0.01 ~ "<0.01",
      between(bp_ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(bp_ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    bp_ftest_cat = factor(bp_ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  ggplot(aes(tau, aic)) +
  geom_line() + 
  geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
  facet_wrap(~population, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "<0.01" = "darkred", 
      "0.01-0.05"  = "red", 
      "0.05-0.10" = "orange", 
      "Not Stat. Sig." = "lightblue"
      )
  ) + 
  geom_vline(
    aes(
      xintercept = bp_breakpoint,
      colour = bp_ftest_cat  
    )
  ) + 
  geom_rect(
    aes(
      xmin = bp_breakpoint - 1.96 * bp_breakpoint_se, 
      xmax = bp_breakpoint + 1.96 * bp_breakpoint_se
    ),
      ymin = -Inf, ymax = Inf,
    alpha = 0.01
  ) + 
  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
  labs(
    title = "Penalised fit (AIC) against proposed changepoint year, various population",
    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
    colour = "Statistical\nsignificance", 
    x = "Year proposed",
    y = "AIC (lower is better)",
    caption = "Vertical lines and bands show estimates and 95% CIs from segmented package"
  )

ggsave("figures/aic_from_gridsearch_and_segmented_bp_estimates.png", height = 20, width = 30, units = "cm", dpi = 150)
```

@@

# Another attempt at a simple summary

It could be we don't get a clear changepoint for some populations because the annual variance as a function of change is particularly large. This would be more an issue for smaller populations than larger populations. The changepoint can therefore be expressed as a standarised unit.

```{r}

mean_sd_chex <- 
  hmd_ex_selected_countries_with_synth %>% 
    filter(code %in% c("DEUTSYNTH", "GBRTENW", "GBR_SCO", "FRATNP", "ITA", "ESP")) %>% 
    filter(between(year, 1979, 2019)) %>% 
    filter(x == 0) %>% 
    group_by(code, sex) %>% 
    arrange(year) %>% 
    mutate(ch_ex = ex - lag(ex)) %>% 
    filter(!is.na(ch_ex)) %>% 
    summarise(
      mean_chex = mean(ch_ex),
      sd_chex   = sd(ch_ex)
    ) %>% 
    mutate(b = mean_chex / sd_chex) %>% 
    ungroup() %>% 
    arrange(desc(mean_chex))

mean_sd_chex

# Now to compare with the changepoint model

get_chpt_coeffs <- function(x){
  tidied_coeffs <- broom::tidy(x)

  tibble(
    mean_chex_before = tidied_coeffs$estimate[1],
    sd_chex_before   = tidied_coeffs$std.error[1],
    mean_chex_after  = tidied_coeffs$estimate[2],
    sd_chex_after    = tidied_coeffs$std.error[2]
  )
}

broke_unbroke <- 
  mean_sd_chex %>% 
    left_join(
      compare_best_breakmodel_with_null %>% 
        ungroup() %>% 
        filter(x == 0) %>% 
        select(code, sex, best_tau, model_alt, ftest_pval) %>% 
        mutate(coeffs = map(model_alt, get_chpt_coeffs)) %>% 
        unnest(coeffs) %>% 
        select(-model_alt) 
    ) %>% 
    mutate(
      ch_as_ratio_of_mean_chex = mean_chex_after / mean_chex,
      pos_chex_after = mean_chex_before + mean_chex_after
    )

# now to visualise in a simple way

expand_grid(
  year = 1980:2019,
  broke_unbroke
  ) %>% 
  mutate(
    broke_val = if_else(year < best_tau, mean_chex_before, mean_chex_before + mean_chex_after)
  ) %>% 
  left_join(
    hmd_ex_selected_countries_with_synth %>% 
      filter(code %in% c("DEUTSYNTH", "GBRTENW", "GBR_SCO", "FRATNP", "ITA", "ESP")) %>% 
      filter(between(year, 1979, 2019)) %>% 
      filter(x == 0) %>% 
      group_by(code, sex) %>% 
      arrange(year) %>% 
      mutate(ch_ex = ex - lag(ex)) %>% 
      filter(!is.na(ch_ex)) %>% 
      ungroup() %>% 
      select(code, year, sex, ch_ex)
  ) %>% 
  ggplot(aes(year)) +
  geom_hline(aes(yintercept = mean_chex)) + 
  geom_hline(yintercept = 0, size = 1.3) + 
  facet_grid(code ~ sex) +
  geom_ribbon(aes(ymin = mean_chex - sd_chex, ymax = mean_chex + sd_chex), alpha = 0.2) + 
  geom_ribbon(aes(ymin = mean_chex - 2 * sd_chex, ymax = mean_chex + 2 * sd_chex), alpha = 0.2) + 
  expand_limits(y = 0) + 
  geom_step(aes(y = broke_val, alpha = 1 - ftest_pval, linetype = ftest_pval < 0.05), colour = "red") + 
  geom_point(aes(y = ch_ex)) +
  scale_alpha_continuous(range = c(0, 1)) + 
  scale_linetype_manual(values = c("FALSE" = "dotted", "TRUE" = "solid")) + 
  guides(alpha = "none", linetype = "none")
